Perturbing SU($n$)-symmetric interactions

### 2020-03-05 collective_sun/sun_perturbation.tex
added section on computing expectation values of multi-local operators

\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}

@@

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
  \label{eq:index_values}
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  In words,
$\C_N\p{g}$ consists of all the ways to choose, for each subset $S$ of
the tensors $\set{w_j}$, a set $k_S$ of $g_S$ distinct spins, with the
restriction that all $k_S$ are orthogonal: if $z\in k_S$, then
$z\notin k_R$ for $R\ne S$.  Having defined the set $\C_N\p{g}$ of
index values, we define the restriction of $k\in\C_N\p{g}$ to values
associated with $w_j$ and $O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
  \label{eq:diagram_factors}
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} { M_j \choose g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  {M_j \choose a_1,a_2,\cdots,a_m}
  \equiv \f{M_j!}{a_1!a_2!\cdots a_m!}
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices as specified by $g$, and the
second product in \eqref{eq:sym_prod_symmetry} accounts for the number
of ways to permute each set of shared indices on every tensor.
Altogether, the symmetry factors $\S\p{g}$ cancel out with the
$\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQPS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

@@

Here we discuss the rules governing symmetry factors that appear when
decomposing diagrams.  In general, eliminating a filled dot from a
region with $F$ filled dots makes a diagram pick up a factor of $1/F$,
e.g.
\begin{align}
  \diagram{example_sym}
  = \f14 \diagram{example_sym_o}
  - \f14 \diagram{example_sym_x}.
\end{align}
This factor can be derived by replacing the sum over $\C_N\p{F}$ for
the filled dots within the region by a sum over $\ZZ_N^{F}$, which
picks up a factor of $1/F!$ to account for the number of ways to
permute $F$ values.  Setting aside one filled dot for replacement by
an empty dot and a cross, the sum over $\ZZ_N^{F-1}$ for the remaining
dots can be changed back to a sum over $\C_N\p{F-1}$, picking up a
factor of $\p{F-1}!$.  The overall factor picked up throughout this
procedure is therefore $\p{F-1}!/F!=1/F$.  Similarly, adding an empty
dot to a region with $E$ empty dots makes the diagram pick up a factor
of $\p{E+1}!/E!=E+1$, so
\begin{align}
  \diagram{example_sym_o}
  = \f23 \diagram{example_sym_oo}
  - \f13 \diagram{example_sym_ox}.
\end{align}
Finally, eliminating a cross by moving a filled dot from an ``old''
region with $F_{\t{old}}$ filled dots into a ``new'' region with
$F_{\t{new}}$ filled dots picks up an overall factor of
$F_{\t{new}}+1$, so
\begin{align}
  \diagram{example_sym_x}
  = 2 \diagram{example_sym_x_elim},
\end{align}
where the factor of $1/F_{\t{old}}$ that is acquired due to removing a
dot from the old region is canceled out by a factor of $F_{\t{old}}$
that accounts for the number of ways to move a dot from the old region
to the new one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Matrix elements of multi-body operators}

\red{TODO: write section}

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}

@@

In order to write out the factors in \eqref{eq:sym_prod_group_start}
explicitly, we identify the set of values that the distinct indices
assigned according to $g$ can take:
\begin{align}
  \C_N\p{g}
  \equiv \set{ k \in \bigotimes_{S\in\PP\p{\ZZ_p}} \C_N\p{g_S}
    : \t{all elements of $k$ are distinct} },
  \label{eq:index_values}
\end{align}
where each $k\in\C_N\p{g}$ has the natural decomposition
$k=\p{k_S:S\in\PP\p{\ZZ_p}}$ with $k_S\in\C_N\p{g_S}$.  In words,
$\C_N\p{g}$ consists of all the ways to choose, for each subset $S$ of
the tensors $\set{w_j}$, a set $k_S$ of $g_S$ distinct spins, with the
restriction that all $k_S$ are orthogonal: if $z\in k_S$, then
$z\notin k_R$ for $R\ne S$.  Having defined the set $\C_N\p{g}$ of
index values, we define the restriction of $k\in\C_N\p{g}$ to values
associated with $w_j$ and $O_j$:
\begin{align}
  k_j \equiv \bigcup_{S\in\PP\p{\ZZ_p} : j\in S} k_S
  = \bigcup_{R\in\PP\p{\ZZ_p\setminus\set{j}}} k_{\set{j}\cup R},
\end{align}
where the union of lists $k_S$ denotes a concatenation of those lists.
These definitions allow us to expand
\begin{align}
  w\p{g} \equiv \sum_{k\in\C_N\p{g}} \prod_{j\in\ZZ_p} w_j\p{k_j},
  &&
  O\p{g} \equiv \prod_{j\in\ZZ_p}
  O_j\p{\ell_j}~\t{for some}~\ell\in\C_N\p{g},
  \label{eq:diagram_factors}
\end{align}
where the choice of $\ell\in\C_N\p{g}$ does not matter after
projection onto permutationally symmetric manifold, and
\begin{align}
  \S\p{g} \equiv \sp{\prod_{j\in\ZZ_p} \C\p{ g_S : j\in S }}
  \sp{\prod_{S\in\PP\p{\ZZ_p}} \p{g_S!}^\abs{S}}
  = \prod_{j\in\ZZ_p} M_j!,
  \label{eq:sym_prod_symmetry}
\end{align}
where the first product in \eqref{eq:sym_prod_symmetry} contains
multinomial coefficients
\begin{align}
  \C\p{a_1,a_2,\cdots,a_m}
  \equiv \f{\p{a_1+a_2+\cdots+a_m}!}{a_1!a_2!\cdots a_m!}
\end{align}
that account for the number of ways to partition the indices of each
$w_j,O_j$ into sets of shared indices as specified by $g$, and the
second product in \eqref{eq:sym_prod_symmetry} accounts for the number
of ways to permute each set of shared indices on every tensor.
Altogether, the symmetry factors $\S\p{g}$ cancel out with the
$\sim M_j!$ prefactors in \eqref{eq:sym_prod_vec}, so
\begin{align}
  \prod_{j\in\ZZ_p} \sum_{k\in\C_N\p{M_j}} w_j\p{k} O_j\p{k}
  \EQPS \sum_{g\in\G\p{\m M}} w\p{g} O\p{g}.
  \label{eq:sym_prod_group}
\end{align}

@@

As seen in \eqref{eq:example_elim_final}, the process of simplifying
diagrams generally involves symmetry factors that need to be kept
track of by hand.  Here we discuss the rules governing symmetry
factors that appear when decomposing diagrams.  In general,
eliminating a filled dot from a region with $F$ filled dots makes a
diagram pick up a factor of $1/F$, e.g.
\begin{align}
  \diagram{example_sym}
  = \f14 \diagram{example_sym_o}
  - \f14 \diagram{example_sym_x}.
\end{align}
This factor can be derived by replacing the sum over $\C_N\p{F}$ for
the filled dots within the region by a sum over $\ZZ_N^{F}$, which
picks up a factor of $1/F!$ to account for the number of ways to
permute $F$ values.  Setting aside one filled dot for replacement by
an empty dot and a cross, the sum over $\ZZ_N^{F-1}$ for the remaining
dots can be changed back to a sum over $\C_N\p{F-1}$, picking up a
factor of $\p{F-1}!$.  The overall factor picked up throughout this
procedure is therefore $\p{F-1}!/F!=1/F$.  Similarly, adding an empty
dot to a region with $E$ empty dots makes the diagram pick up a factor
of $\p{E+1}!/E!=E+1$, so
\begin{align}
  \diagram{example_sym_o}
  = \f23 \diagram{example_sym_oo}
  - \f13 \diagram{example_sym_ox}.
\end{align}
Finally, eliminating a cross by moving a filled dot from an ``old''
region with $F_{\t{old}}$ filled dots into a ``new'' region with
$F_{\t{new}}$ filled dots picks up an overall factor of
$F_{\t{new}}+1$, so
\begin{align}
  \diagram{example_sym_x}
  = 2 \diagram{example_sym_x_elim},
\end{align}
where the factor of $1/F_{\t{old}}$ that is acquired due to removing a
dot from the old region is canceled out by a factor of $F_{\t{old}}$
that accounts for the number of ways to move a dot from the old region
to the new one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Matrix elements of individual multi-spin operators}

The process of constructing and simplifying diagrams to compute the
coefficients $w\p{g}$ of the diagrammatic operator product expansion
in \eqref{eq:sym_prod_group} is systematic enough to be converted into
an algorithm for execution on a computer.  The last step necessary to
automate the evaluation of \eqref{eq:sym_prod_group} is expansion of
the multi-spin operators $O\p{g}$ in a basis for the permutationally
symmetric manifold $\M_0$.  Without loss of generality, we consider a
single permutationally symmetric $M$-spin operator $Q$; our task is
essentially to find coefficients of the expansion
\begin{align}
  \P_0 Q \P_0
  = \sum_{a,b} \bk{a|Q|b} \op{a}{b},
\end{align}
where $\set{\ket{a}},\set{\ket{b}}$ are complete bases for $\M_0$, and
the choice of $M$ spins on which $Q$ acts is arbitrary.  In a system
of $N$ spins with $n$ states each, a suitable basis for $\M_0$ can be
labeled by the occupation number of each single-spin state, namely
$a=\p{a_1,a_2,\cdots,a_n}$ with each $a_j\ge 0$ and $\sum_ja_j=N$.
Written out explicitly,
\begin{align}
  \ket{a} = \f1{\sqrt{\C\p{a}}}
  \sum_{\substack{\t{distinct}\\\t{permutations}\\\Pi~\t{of}~\tilde a}}
  \Pi \ket{\tilde a},
  &&
  \C\p{a} \equiv \f{\p{\sum_j a_j}!}{\prod_ja_j!}
  &&
  \ket{\tilde a} \equiv \bigotimes_{j=1}^n \ket{j}^{\otimes a_j},
\end{align}
where the multinomial coefficient $\C\p{a}$ counts the number of
distinct ways to permute the tensor factors of $\ket{\tilde a}$, and
enforces $\bk{a|a}=1$.  We denote the set of all valid assignments of
$S$ spins to $n$ states by $\A_n\p{S}$, such that the standard basis
for the permutationally symmetric manifold of $S$ spins is
$\set{\ket{a}:a\in\A_n\p{S}}$.  Identifying this standard basis allows
us to expand
\begin{align}
  \bk{a|Q|b} = \sum_{\alpha,\beta\in\A_n\p{M}}
  \delta_{a-\alpha,b-\beta}
  \sqrt{\C\p{a-\alpha}\C\p{b-\beta}}
  \sqrt{\f{\C\p{\alpha}\C\p{\beta}}{\C\p{a}\C\p{b}}}
  \bk{\alpha|Q|\beta},
  \label{eq:multi_body_eval}
\end{align}
where the differences $a-\alpha$ and $b-\beta$ are understood to be
taken element-wise; $\delta_{a-\alpha,b-\beta}=1$ if
$a-\alpha=b-\beta$ and zero otherwise; and for ease of notation we
define $\C\p{c}\equiv 0$ if any entries of $c$ are negative.  We sum
over both $\alpha$ and $\beta$ above merely to keep the expression
symmetric with respect to transposition; in practice, one can simply
sum over $\alpha\in\A_n\p{M}$ and set $\beta\equiv b-a+\alpha$.  Note
that, by slight abuse of notation, the operator $Q$ on the left of
\eqref{eq:multi_body_eval} acts on an arbitrary choice of $M$ spins
(out of $N$), whereas the operator $Q$ on the right of
\eqref{eq:multi_body_eval} is simply an $M$-spin operator, with matrix
elements $\bk{\alpha|Q|\beta}$ evaluated with respect to the $M$-spin
states $\ket\alpha,\ket\beta$.
